---
title: "padmaska_m5_as3"
author: "Artsiom Padmaska"
date: "2024-01-23"
output: html_document
---


```{r, results='hide', echo=FALSE, warning=FALSE, message=FALSE}
library(AER)
library(MASS)
library(lmtest)
library(sandwich)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(GGally)
library(readxl)
theme_set(theme_bw())

mycol <- rgb(1,0,0,0.1,maxColorValue=1)

seismic_data <- read_excel("seismic.xlsx", sheet = "data")

head(seismic_data)

```
```{r, echo=FALSE, fig.height=15, fig.width=15, warning=F, message=F}
seismic_data %>% 
  mutate_at(c("seismic", "seismoacoustic", "shift", "ghazard", "class"), as.factor) %>%
  ggpairs(mapping = aes(alpha=0.5, fill='red'))

```
diagonal cells for nbumps >=6 don't have nothing, let's check if they have anything but 0
```{r}
unique(seismic_data$nbumps6)
unique(seismic_data$nbumps7)
unique(seismic_data$nbumps89)
```
There are only 0-s, so we need to delete this features. Also, let's check if the binary + nbumps features balanced?
```{r, echo=FALSE}
seismic_data = subset(seismic_data, select = -c(nbumps6,nbumps7,nbumps89) )
```

```{r}
ftable(seismic_data$seismic, seismic_data$class)
ftable(seismic_data$seismoacoustic, seismic_data$class)
ftable(seismic_data$shift, seismic_data$class)
ftable(seismic_data$ghazard, seismic_data$class)
ftable(seismic_data$ghazard, seismic_data$nbumps)
ftable(seismic_data$ghazard, seismic_data$nbumps2)
ftable(seismic_data$ghazard, seismic_data$nbumps3)
ftable(seismic_data$ghazard, seismic_data$nbumps4)
ftable(seismic_data$ghazard, seismic_data$nbumps5)
```
Only 48 result of shift obtained by the seismoacoustic method and 30  based on registration coming form GMax only are high hazard. We won't be able to reliably estimate a coefficient of this feature, so I'll make result c be result b.

```{r}
seismic_data$seismoacoustic[seismic_data$seismoacoustic == "c"] <- "b"
seismic_data$ghazard[seismic_data$ghazard == "c"] <- "b"
```

Standard Poisson model:
```{r, echo=FALSE}
m1 <- glm(class~., family=poisson(), data=seismic_data)
summary(m1)
```
```{r, echo=FALSE}
m0 <- glm.nb(class~., data=seismic_data)
summary(m0)
```
```{r}
1 - m1$deviance / m1$null.deviance
```
```{r}
1 - m0$deviance / m0$null.deviance
```
```{r, echo=FALSE}
m2 <- glm(class~shift + gpuls + nbumps + nbumps2 + nbumps3 + nbumps4, family=poisson(), data=seismic_data)
summary(m2)
```
Testing the hypothesis of equidispersion against the alternative of overdispersion:
```{r, echo=FALSE}
dispersiontest(m1, alternative="greater")
```
p-value > 0.05, so we can reject overdispersion 

```{r}
1 - m2$deviance / m2$null.deviance
```
```{r}
lrtest(m1,m2)
```
no significant difference, so m2 is not worse than m1

```{r}
add1(m2, ~ . + seismic + seismoacoustic + genergy + gdenergy + gdpuls + ghazard + nbumps5 + energy + maxenergy, test="LRT")
```


```{r, echo=FALSE}
m3 <- glm(class~shift + gpuls + nbumps + nbumps2 + nbumps3 + nbumps4 + nbumps5, family=poisson(), data=seismic_data)
summary(m3)
```

```{r}
1 - m3$deviance / m3$null.deviance
```
```{r}
lrtest(m3,m2)
```
model 3 is significantly better than model 2

```{r}
lrtest(m1,m3)
```
model 3 is not worse than model 1


```{r}
add1(m3, ~ . + seismic + seismoacoustic + genergy + gdenergy + gdpuls + ghazard + energy + maxenergy, test="LRT")
```

None provides a significant improvement.
```{r, echo=FALSE}
add1(m3, ~ .^2, test="LRT")

```

```{r, echo=FALSE}
m4 <- glm(class~shift + gpuls + nbumps + nbumps2 + nbumps3 + nbumps4 + nbumps5 + shift*nbumps + shift*nbumps3 + shift*nbumps4 + gpuls*nbumps + nbumps*nbumps2 + nbumps*nbumps3, family=poisson(), data=seismic_data)
summary(m4)
lrtest(m3,m4)
```
$R^2_{DEV}$
```{r}
1 - m4$deviance / m4$null.deviance
```


```{r, echo=FALSE}
add1(m4, ~ .^2, test="LRT")
```
No more significant improvements.


```{r, echo=FALSE}
summary(m4)
cov.robust     <- vcovHC(m4, type="HC0")
se.robust      <- sqrt(diag(cov.robust))
coeffs         <- coef(m4)
t.robust       <- coeffs / se.robust
summary.robust <- cbind(coeffs, se.robust, t.robust, pvalue = 2*(1-pnorm(abs(coeffs/se.robust))))
print(summary.robust)
```


No significant difference, we can trust our model

```{r}
yhat <- predict(m4, type="response")
plot(yhat, cooks.distance(m4), pch=20, xlab="Estimated conditional mean", ylab="Cook's distance")
lines(c(0,10), c(0.03,0.03), col="red", lwd=2)
```


```{r}
seismic_data2 <- seismic_data[cooks.distance(m4)<0.03,]
m5 <- glm(class~shift + gpuls + nbumps + nbumps2 + nbumps3 + nbumps4 + nbumps5 + shift*nbumps + shift*nbumps3 + shift*nbumps4 + gpuls*nbumps + nbumps*nbumps2 + nbumps*nbumps3, family=poisson(), data=seismic_data2)
summary(m5)
```

```{r}
tmp <- cbind(coefficients(m4), coefficients(m5))
colnames(tmp) <- c("All data", "Filtered by leverage")
tmp
```

```{r, echo=FALSE}
yhat <- predict(m4, type="response")
plot(jitter(seismic_data$class), yhat, col=mycol, pch=20, ylab="Estimated decision attribute", xlab="decision attribute")
```




The exponents of the coefficients of the final model and confidence intervals for them:
```{r, echo=FALSE, message=FALSE}
exp(coefficients(m4))[-1]
exp(confint(m4))[-1,]
```


Conclusions:

Shift Type: Coal-getting shifts (W) have a higher risk (`r round(exp(coefficients(m4))["shiftW"],2)` times higher) of high energy seismic bumps compared to preparation shifts. (`r round(exp(confint(m4))["shiftW",], 2)`).

Seismic Activity: 

Each additional pulse recorded by GMax increases the odds of a high energy seismic bump by a factor of `r round(exp(coefficients(m4))["gpuls"],4)`, with a very narrow confidence interval (`r round(exp(confint(m4))["gpuls",], 4)`), suggesting a very slight effect.

Each additional seismic bump recorded increases the odds of a high energy seismic bump by a factor of `r round(exp(coefficients(m4))["nbumps"],2)`. However, the confidence interval is very wide (`r round(exp(confint(m4))["nbumps",], 2)`), indicating considerable uncertainty in this estimate.

Energy Range of Bumps: The impact of seismic bumps varies across different energy ranges, with higher energy bumps (nbumps5) having a much lower odds ratio. But the confidence intervals vary widely, suggesting differing levels of uncertainty.

Interactions: 

The interaction between coal-getting shifts and the number of seismic bumps decreases the odds of a high energy seismic bump by `r (1 - round(exp(coefficients(m4))["shiftW:nbumps"],2)) * 100`% (CI: `r round(exp(confint(m4))["shiftW:nbumps",], 2)`), indicating a mitigating effect of coal-getting shifts on seismic risks.



For seismic bumps in the energy range $[10^3,10^4)$, the interaction with coal-getting shifts reduces the odds of a high energy seismic bump by `r (1 - round(exp(coefficients(m4))["shiftW:nbumps3"],2)) * 100`% (CI: `r round(exp(confint(m4))["shiftW:nbumps3",], 2)`), suggesting effective risk controls in these shifts for this energy range.

The interaction of coal-getting shifts with seismic bumps in the energy range $[10^4,10^5)$ significantly lowers the odds of a high energy seismic bump by `r (1 - round(exp(coefficients(m4))["shiftW:nbumps4"],2)) * 100`% (CI: `r round(exp(confint(m4))["shiftW:nbumps4",], 2)`), pointing to strong safety measures in these shifts for higher energy bumps.


The combined effect of the number of pulses and seismic bumps on the occurrence of a high energy seismic bump is almost negligible (odds ratio: `r round(exp(coefficients(m4))["gpuls:nbumps"],4)`; CI: `r round(exp(confint(m4))["gpuls:nbumps",], 4)`), indicating the number of pulses does not significantly influence seismic risks when combined with the number of bumps.


The interaction between the total number of seismic bumps and the number of seismic bumps in the energy range $[10^2,10^3)$ decreases the odds of a high energy seismic bump by `r (1 - round(exp(coefficients(m4))["nbumps:nbumps2"],2)) * 100`% (CI: `r round(exp(confint(m4))["nbumps:nbumps2",], 2)`). This suggests a slight mitigating effect when these specific ranges of seismic bumps occur together.

The interaction between the total number of seismic bumps and the number of seismic bumps in the energy range $[10^3,10^4)$ decreases the odds of a high energy seismic bump by `r (1 - round(exp(coefficients(m4))["nbumps:nbumps3"],2)) * 100`% (CI: `r round(exp(confint(m4))["nbumps:nbumps3",], 2)`). This indicates a small, potentially negligible combined effect for these particular ranges of seismic activity. 













